\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-02-04 paper.tex
Tried 5-fold cross validation to compute mse and investigate the distribution (not specific change with cross validation)

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

Typically, the MSE plot will have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope will gradually decline or level off, indicating
that a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} (b) has created a
graph that shows how the Mean Squared Error (MSE) changes depending on
the number of bins assigned in the hexagonal grid for various NLDR
techniques applied to the S-curve data set. As per the rule of thumb
mentioned earlier, the effective number of bins for each NLDR method is
shown in the \textbf{?@tbl-msebinsnldr}. Furthermore,
Figure~\ref{fig-modelScurve} shows the 2D models constructed for each
NLDR technique with the selected total number of bins and how the models
look in high-D with original data.

@@

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
Figure~\ref{fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in Figure~\ref{fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

After considering the models constructed from the algorithm for each
NLDR techniques and visualizing them in high-D space with the original
data, shows that all the NLDR techniques capture the non-linearity of
the S-curve (see videos of the models linked in
Figure~\ref{fig-modelScurve}). But thorough examination of each model in
high-D space shows some surprising findings. The model of PHATE is
flattened and lack of capturing the width of S-curve (see videos of the
model with PHATE linked in Figure~\ref{fig-modelScurve}). This nature of
PHATE is also evicted by the considerably higher AIC and MSE values for
PHATE as shown in Figure~\ref{fig-diagnosticpltScurve}. Furthermore,
there are two edges that are not connected to any points in the model of
TriMAP (see see videos of the model with TriMAP linked in
Figure~\ref{fig-modelScurve}). This model shows how the patterns in 2D
representation of TriMAP capture by the model (see
Figure~\ref{fig-nldervis} (d)). As shown in Figure~\ref{fig-nldervis}
(b), UMAP have four clusters in a curvilinear structure. However, the
model reveals that even though UMAP shows a curvilinear in 2D, UMAP
actually capture the structure of the S-curve without any disjoint in
the model (see see videos of the model with UMAP linked in
Figure~\ref{fig-modelScurve}). In the 2D layouts of tSNE and PaCMAP (see
Figure~\ref{fig-nldervis} (a), (e)), there are some sparse areas that
are captured by the models effectively (see see videos of the model with
tSNE and PaCMAP linked in Figure~\ref{fig-modelScurve}).

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-modelScurve-1.pdf}

\caption{\label{fig-modelScurve}Is there a good model to represent the
original data in 2D space or are they all providing equivalent
information?, (a) Model in the 2D space with tSNE
(\url{https://youtu.be/uy9SkAo6gAE}), (b) Model in the 2D space with
UMAP (\url{https://youtu.be/0MJDhHrh_Ug}), (c) Model in the 2D space
with PHATE (\url{https://youtu.be/HbVv0uy0QWk}), (d) Model in the 2D
space with TriMAP (\url{https://youtu.be/2OGwipAjzc8}), and (e) Model in
the 2D space with PaCMAP (\url{https://youtu.be/pkokI8d-cBk}).}

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, breakable, frame hidden, enhanced, sharp corners]}{\end{tcolorbox}}\fi

@@

Typically, the MSE plot will have a steep slope at the beginning,
indicating that a smaller number of bins causes a larger amount of
error. Then, the slope will gradually decline or level off, indicating
that a higher number of bins generates a smaller error. This MSE plot is
useful in determining the effective number of bins required to construct
the 2D model. Figure~\ref{fig-diagnosticpltScurve} (b) has created a
graph that shows how the Mean Squared Error (MSE) changes depending on
the number of bins assigned in the hexagonal grid for various NLDR
techniques applied to the S-curve data set. As per the rule of thumb
mentioned earlier, the effective number of bins for each NLDR method is
shown in the \textbf{?@tbl-msebinsnldr}. Furthermore,
\textbf{?@fig-modelScurve} shows the 2D models constructed for each NLDR
technique with the selected total number of bins and how the models look
in high-D with original data.

@@

Consequently, it is not necessary to remove low-density hexagons when
constructing the 2D model for the S-curve dataset in NLDR techniques,
except for PHATE. While there may be some fluctuations in MSE between
benchmark values of 0 to 1, the absence of significant changes suggests
that maintaining low-density hexagons is preferable. However, to choose
the best selection, it's recommended to examine the benchmark values
around the first local minimum in the MSE curve. However, the MSE for
PHATE fluctuates when benchmark values range between 0 and 1. Therefore,
it is necessary to explore the benchmark values surrounding the first
local minimum to determine an effective benchmark value. There is no
universal rule for selecting a process. As shown in
Figure~\ref{fig-diagnosticpltScurvelwd}, the MSE values for PHATE
fluctuate, and the first local minimum is at \(0\). Therefore, nearby
benchmark values such as \(0.01\), \(0.04\), and \(0.06\) were explored
to identify the best value for preserving the data structure. In this
case, a benchmark value of \(0.06\) was chosen. The resulting 2D model
is shown in \textbf{?@fig-modelScurve} (c), while the video link
provides a visual representation of how the model appears in high-D.

@@

After considering the models constructed from the algorithm for each
NLDR techniques and visualizing them in high-D space with the original
data, shows that all the NLDR techniques capture the non-linearity of
the S-curve (see videos of the models linked in
\textbf{?@fig-modelScurve}). But thorough examination of each model in
high-D space shows some surprising findings. The model of PHATE is
flattened and lack of capturing the width of S-curve (see videos of the
model with PHATE linked in \textbf{?@fig-modelScurve}). This nature of
PHATE is also evicted by the considerably higher AIC and MSE values for
PHATE as shown in Figure~\ref{fig-diagnosticpltScurve}. Furthermore,
there are two edges that are not connected to any points in the model of
TriMAP (see see videos of the model with TriMAP linked in
\textbf{?@fig-modelScurve}). This model shows how the patterns in 2D
representation of TriMAP capture by the model (see
Figure~\ref{fig-nldervis} (d)). As shown in Figure~\ref{fig-nldervis}
(b), UMAP have four clusters in a curvilinear structure. However, the
model reveals that even though UMAP shows a curvilinear in 2D, UMAP
actually capture the structure of the S-curve without any disjoint in
the model (see see videos of the model with UMAP linked in
\textbf{?@fig-modelScurve}). In the 2D layouts of tSNE and PaCMAP (see
Figure~\ref{fig-nldervis} (a), (e)), there are some sparse areas that
are captured by the models effectively (see see videos of the model with
tSNE and PaCMAP linked in \textbf{?@fig-modelScurve}).
